Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-11-13 grove-paper.tex
Revisions to Section 2.3 (Michael D. Adams)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureMultiUserCycles
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{MultiCycleTimes}{L}
\newvertex{MultiCycleTimes}{*}
\newedge{MultiCyclePlus}{R}
\newvertex{MultiCyclePlus}{+}
\newedge{MultiCycleAliceTimes}{R}
\newedge{MultiCycleAlicePlus}{L}
\newedge{MultiCycleBobPlus}{R}
\newedge{MultiCycleBobTimes}{L}
\newcommand{\figureMultiUserCycles}{
\begin{figure*}
%\hfill
\hskip0.12\linewidth
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (fig-Multi-parent-e-x) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:Multi-parent-e}}}};
%%%%
\path (0cm,0cm) node (fig-Multi-User Cycles-a) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-parent-e-x) to[star] (-2em, 0cm |- fig-Multi-User Cycles-a);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-a}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
    "\vMultiCycleTimes" [> "\eMultiCycleTimes"', >alice edge, alice node],
    "\vMultiCyclePlus" [> "\eMultiCyclePlus", >alice edge, alice node]
   },
   {}
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Multi-User Cycles-b) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-User Cycles-a) to[star] (-2em, 0cm |- fig-Multi-User Cycles-b);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-b}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
   },
   {
    "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes", >alice edge, alice node] -> {
     "\vMultiCyclePlus" [> "\eMultiCycleAlicePlus"', >alice edge, alice node],
     {}
    }
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Multi-User Cycles-c) [anchor=base] {\strut};
\path [draw,->,bob step] (fig-Multi-User Cycles-a) to[out=15,in=165,star] (-2em, 0cm |- fig-Multi-User Cycles-c);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-c}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
   },
   {
    "\vMultiCyclePlus" [> "\eMultiCycleBobPlus", >bob edge, bob node] -> {
     "\vMultiCycleTimes" [> "\eMultiCycleBobTimes"', >bob edge, bob node],
     {}
    }
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Multi-User Cycles-d) [anchor=base] {\strut};
\path [draw,->,merge step] (fig-Multi-User Cycles-b) to[out=15,in=165] (-2em, 0cm |- fig-Multi-User Cycles-d);
\path [draw,->,merge step] (fig-Multi-User Cycles-c) to (-2em, 0cm |- fig-Multi-User Cycles-d);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-d}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'anchor=-15] -> {
   },
   v1/"\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"anchor=155, >merge edge, merge node],
   {},
   v2/"\vMultiCyclePlus" [> "\eMultiCycleBobPlus"anchor=-165, >merge edge, merge node]
  }
 }
};
\path[draw,-{>[bend]},merge edge] (v1) to ["\eMultiCycleAlicePlus"anchor=south,out=-60,in=-120] (v2);
\path[draw,-{>[bend]},merge edge] (v2) to ["\eMultiCycleBobTimes"anchor=north,out=-90,in=-90] (v1);
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Example of Cycles.}
\label{fig:Multi-User Cycles}
\end{figure*}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureDisconnection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{DisconnectionAlice}{R}
\newedge{DisconnectionBob}{L}
\newcommand{\figureDisconnection}{
\begin{figure*}
%\hfill
\hskip0.12\linewidth
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (fig-Multi-User Cycles-d-x) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:Multi-User Cycles-d}}}};
%%%%
\path (0cm,0cm) node (fig-Disconnection-a) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-User Cycles-d-x) to[star] (-2em, 0cm |- fig-Disconnection-a);
\end{tikzpicture}
}
\label{fig:Disconnection-a}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'],
   "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-b) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Disconnection-a) to[star] (-2em, 0cm |- fig-Disconnection-b);
\end{tikzpicture}
}
\label{fig:Disconnection-b}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
    {},
    "\vMultiCycleTimes" [> "\eDisconnectionAlice", >alice edge]
   },
   {}
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-c) [anchor=base] {\strut};
\path [draw,->,bob step] (fig-Disconnection-a) to[out=15,in=165,star] (-2em, 0cm |- fig-Disconnection-c);
\end{tikzpicture}
}
\label{fig:Disconnection-c}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   {},
   "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"] -> {
    "\vSimpleTimes" [> "\eDisconnectionBob"', >bob edge],
    {}
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-d) [anchor=base] {\strut};
\path [draw,->,merge step] (fig-Disconnection-b) to[out=15,in=165] (-2em, 0cm |- fig-Disconnection-d);
\path [draw,->,merge step] (fig-Disconnection-c) to (-2em, 0cm |- fig-Disconnection-d);
\end{tikzpicture}
}
\label{fig:Disconnection-d}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"]
 }
};
\path (-0.5cm,-2cm) graph[graph style] { plus/"\vSimpleTimes" };
\path ( 0.5cm,-2cm) graph[graph style] { times/"\vMultiCycleTimes" };
\path[draw,->,merge edge] (times) to ["\eMultiCycleBobTimes"'anchor=north,out=-135,in=-45] (plus);
\path[draw,->,merge edge] (plus) to ["\eMultiCycleBobPlus"anchor=south,out=45,in=135] (times);
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Example of Disconnection.}
\label{fig:Disconnection}
\end{figure*}
}

@@

Having described single-user actions,
we turn our attention to how Grove behaves in the presence of multiple users. 
Everything we discuss in this section generalizes to any number of users,
but for simplicity we consider only two: Alice and Bob. 
Alice and Bob each maintain their own editor instance, and perform 
actions relative to its state (using their own cursors, which we discuss in \autoref{sub:Cursors}).
We start each user where we left off in the previous section, i.e., in the state shown in \autoref{fig:Move-b}.

@@

We assume multiple instances

Common scenarios first and then consider more unusual scenarios
that cause cycles or multiple parents

@@

To start, Alice and Bob are synchronized with each other and
both have the graph in \autoref{fig:Move-b}, which
represents the code~\texttt{\hole{} + (\hole{} + y)}.

Alice adds~\vDifferentAlice{} as the left child of \vSimpleTimes{}
and Bob changes \vConstructionY{} to \vDifferentBob{}.
Before transmitting their edits to each other,
Alice thus has the graph in \autoref{fig:Editing Different Parts of the Code-a}
and Bob has the graph in \autoref{fig:Editing Different Parts of the Code-b}.

@@

Note that aside from transmitting the edge deletions and additions,
the merge requires no coordination between Alice and Bob.
For example, suppose that Alice and Bob did not transmit their edits to each other
but to a third user, Chris, that has the original graph in \autoref{fig:Move-b}.
Chris can apply the edits from Alice and Bob in any order and will always get
the graph in \autoref{fig:Editing Different Parts of the Code-c}.

Stated more formally, our edit model obeys the commutativity shown in \autoref{fig:Commutativity}.
For any two sequences of edits $\vec{\alpha}$ and $\vec{\alpha}^\prime$,
applying them in either order produces the same results.

@@

After the edits in the previous subsection synchronize,
Alice changes~\vDifferentAlice{} to~\vNestedAlice{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-a}.
On the other hand, Bob moves~\vSimpleTimes{} and its children from the left child of~\vWrapPlus{} to its right child
by deleting~\eMoveTimes{} and adding~\eNestedBob{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-b}.
Since vertex and edge identifiers are preserved by these,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the graph in \autoref{fig:Editing Nested Parts of the Code-c}.

A consequence of this design is that we allow users to modify
parts of the code that have been deleted.  For example,
Alice might receive Bob's deletion of~\eMoveTimes{} before making her
changes and before receiving Bob's addition of~\eNestedBob{}.
Her edits would then be within the code deleted by Bob.
TODO: She can restore that code using a restoration action add a fresh edge to it.

@@

TODO: Multi-user merges can still create multi-child conflicts,
but it is better to not interpret user actions in a way that creates them.
How to handle multiple children in the same position is explored in \autoref{sub:Multi-child conflicts}.

Finally, we address our first kind of conflict between edits.
Alice adds~\vMultiChildAlice{} as the right child of~\vWrapPlus{}
while Bob adds~\vMultiChildBob{} also as the right child of~\vWrapPlus{}.
These result in the graphs in \autoref{fig:Multi-child-a} and \autoref{fig:Multi-child-b}, respectively.
In particular pay attention to the edges going out of the \texttt{R} position of \vWrapPlus{}.
In \autoref{fig:Multi-child-a} it is~\eMultiChildAlice{}, which points to~\vMultiChildAlice{},
and in \autoref{fig:Multi-child-b} it is~\eMultiChildBob{}, which points to~\vMultiChildBob{}.

We merge these by including \emph{both} edges as edges going out of the \texttt{R} position
of \vWrapPlus{}.
This results in \autoref{fig:Multi-child-c}.
The conflict between Alice's and Bob's changes is
represented by the fact that there are multiple edges going out of one index (i.e., the \texttt{R}
position of \vWrapPlus{}).
In the graph representation, we flag this conflict by highlighting its edges
in red (TODO).
In the textual representation, we flag this conflict with the notation~\texttt{\{ x | y \}}.
This is a conflict that must eventually be resolved by the user by deleting
one of the edges and editing the other to be the merged value.
For example, one could delete \eMultiChildBob{} and then change~\vMultiChildAlice{}
to be~\texttt{x * y}.
Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code is run.\footnote{It
may be possible to develop evaluation models that allow these sorts of conflicts,
but that is beyond the scope of this paper.}

TODO: need example of fully-automatic resolution and semi-automatic resolution

Multiple adds cause conflicts
(Not usually presented in the UI)

@@

Our tree-based model is similar in that it presents alternate versions
of the code, and it is up to the user to replace those alternates with the desired merged result.
Our model is different in that merge conflicts in the line-based model
prevent some semantic editor services from running
that could continue to run in the tree-based model because
code inside and outside the conflict is still valid code.
It is just the point where they meet the poses a potential problem for editor services.
Another place our model is different
is that changes are explicit rather than being inferred.
Finally, in our model, since all changes are in terms of
the structure of the code rather than lines,
all changes respect that structure.
This is unlike the line-based model where there is no guarantee
that a merge conflict follows the structure of the code.
Finally, our model differs from the traditional model
in that we have explict add and remove actions
where the traditional model infers these from a differencing algorithm.

Note that as a convenience to the user, certain simple
conflicts might be automatically resolved,
but we consider this a higher-level, user-interface consideration.

@@

In addition to conflicts due to multiple edges coming \emph{out of} child position of a particular vertex,
conflicts can manifest as multiple edges going \emph{into} a particular vertex.
We call these \emph{multi-parent conflicts}.

An example of this is shown in \autoref{fig:Multi-parent}.
\autoref{fig:Multi-parent-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(w + \hole{}) + \hole{})}.
Alice moves the variable~\texttt{w} to the left-most hole,
while Bob moves it to the right-most hole.
This results in the graphs in \autoref{fig:Multi-parent-b} and \autoref{fig:Multi-parent-c},
which represent respectively the expression~\texttt{(\hole{} + w) + \hole{}} for Alice
and the expression~\texttt{(\hole{} + \hole{}) + w} for Bob.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes~\eNestedAlice{} and adds~\eMultiParentAlice{}, while
Bob deletes~\eNestedAlice{} and adds~\eMultiParentBob{}.
As explained in Subsection~TODO:REF, deletes are idempotent so Alice and Bob both deleting~\eNestedAlice{} is not a problem.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob) point to the same vertex.
This generates \autoref{fig:Multi-parent-d}, in which~\vNestedAlice{} has two parents pointing to it.
We call this situation a multi-parent conflict.

@@

TODO: NEEDS UI MOCKUP

TODO: paragraph about why single-user cycles cannot happen

While single-user edits can be checked to see
whether they create a cycle and thus the system
can immediately warn the user,
multi-user edits may not manifest cycles until after
the edits from multiple users are merged.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureMultiUserCycles
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{MultiCycleTimes}{L}
\newvertex{MultiCycleTimes}{*}
\newedge{MultiCyclePlus}{R}
\newvertex{MultiCyclePlus}{+}
\newedge{MultiCycleAliceTimes}{R}
\newedge{MultiCycleAlicePlus}{L}
\newedge{MultiCycleBobPlus}{R}
\newedge{MultiCycleBobTimes}{L}
\newcommand{\figureMultiUserCycles}{
\begin{figure*}
%\hfill
\hskip0.12\linewidth
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (fig-Multi-parent-e-x) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:Multi-parent-e}}}};
%%%%
\path (0cm,0cm) node (fig-Multi-User Cycles-a) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-parent-e-x) to[star] (-2em, 0cm |- fig-Multi-User Cycles-a);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-a}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
    "\vMultiCycleTimes" [> "\eMultiCycleTimes"', >alice edge, alice node],
    "\vMultiCyclePlus" [> "\eMultiCyclePlus", >alice edge, alice node]
   },
   {}
  }
 }
};
\end{tikzpicture}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Multi-User Cycles-b) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-User Cycles-a) to[star] (-2em, 0cm |- fig-Multi-User Cycles-b);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-b}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
   },
   {
    "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes", >alice edge, alice node] -> {
     "\vMultiCyclePlus" [> "\eMultiCycleAlicePlus"', >alice edge, alice node],
     {}
    }
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Multi-User Cycles-c) [anchor=base] {\strut};
\path [draw,->,bob step] (fig-Multi-User Cycles-a) to[out=15,in=165,star] (-2em, 0cm |- fig-Multi-User Cycles-c);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-c}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
   },
   {
    "\vMultiCyclePlus" [> "\eMultiCycleBobPlus", >bob edge, bob node] -> {
     "\vMultiCycleTimes" [> "\eMultiCycleBobTimes"', >bob edge, bob node],
     {}
    }
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Multi-User Cycles-d) [anchor=base] {\strut};
\path [draw,->,merge step] (fig-Multi-User Cycles-b) to[out=15,in=165] (-2em, 0cm |- fig-Multi-User Cycles-d);
\path [draw,->,merge step] (fig-Multi-User Cycles-c) to (-2em, 0cm |- fig-Multi-User Cycles-d);
\end{tikzpicture}
}
\label{fig:Multi-User Cycles-d}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'anchor=-15] -> {
   },
   v1/"\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"anchor=155, >merge edge, merge node],
   {},
   v2/"\vMultiCyclePlus" [> "\eMultiCycleBobPlus"anchor=-165, >merge edge, merge node]
  }
 }
};
\path[draw,-{>[bend]},merge edge] (v1) to ["\eMultiCycleAlicePlus"anchor=south,out=-60,in=-120] (v2);
\path[draw,-{>[bend]},merge edge] (v2) to ["\eMultiCycleBobTimes"anchor=north,out=-90,in=-90] (v1);
\end{tikzpicture}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.17\linewidth}
\centering
\caption{}
\label{fig:Multi-User Cycles-e}
TODO: UI
\end{subfigure}
\hfill{}
\caption{Example of Cycles.}
\label{fig:Multi-User Cycles}
\end{figure*}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureDisconnection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{DisconnectionAlice}{R}
\newedge{DisconnectionBob}{L}
\newcommand{\figureDisconnection}{
\begin{figure*}
%\hfill
\hskip0.12\linewidth
%%%%%%%%
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (fig-Multi-User Cycles-d-x) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:Multi-User Cycles-d}}}};
%%%%
\path (0cm,0cm) node (fig-Disconnection-a) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Multi-User Cycles-d-x) to[star] (-2em, 0cm |- fig-Disconnection-a);
\end{tikzpicture}
}
\label{fig:Disconnection-a}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'],
   "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-b) [anchor=base] {\strut};
\path [draw,->,alice step] (fig-Disconnection-a) to[star] (-2em, 0cm |- fig-Disconnection-b);
\end{tikzpicture}
}
\label{fig:Disconnection-b}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedBob"'] -> {
    {},
    "\vMultiCycleTimes" [> "\eDisconnectionAlice", >alice edge]
   },
   {}
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-c) [anchor=base] {\strut};
\path [draw,->,bob step] (fig-Disconnection-a) to[out=15,in=165,star] (-2em, 0cm |- fig-Disconnection-c);
\end{tikzpicture}
}
\label{fig:Disconnection-c}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   {},
   "\vMultiCycleTimes" [> "\eMultiCycleAliceTimes"] -> {
    "\vSimpleTimes" [> "\eDisconnectionBob"', >bob edge],
    {}
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.17\textwidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (fig-Disconnection-d) [anchor=base] {\strut};
\path [draw,->,merge step] (fig-Disconnection-b) to[out=15,in=165] (-2em, 0cm |- fig-Disconnection-d);
\path [draw,->,merge step] (fig-Disconnection-c) to (-2em, 0cm |- fig-Disconnection-d);
\end{tikzpicture}
}
\label{fig:Disconnection-d}
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"]
 }
};
\path (-0.5cm,-2cm) graph[graph style] { plus/"\vSimpleTimes" };
\path ( 0.5cm,-2cm) graph[graph style] { times/"\vMultiCycleTimes" };
\path[draw,->,merge edge] (times) to ["\eMultiCycleBobTimes"'anchor=north,out=-135,in=-45] (plus);
\path[draw,->,merge edge] (plus) to ["\eMultiCycleBobPlus"anchor=south,out=45,in=135] (times);
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.17\linewidth}
\centering
\caption{}
\label{fig:Disconnection-e}
TODO: UI
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Example of Disconnection.}
\label{fig:Disconnection}
\end{figure*}
}

@@

Having described single-user actions,
we turn our attention to how Grove handles multiple users. 
This section generalizes to any number of users,
but for simplicity we consider only two: Alice and Bob. 
Alice and Bob each maintain their own editor instance and perform 
actions relative to its state (using their own cursors, which we discuss in \autoref{sub:Cursors}).
We start each user where we left off in the previous section with the state shown in \autoref{fig:Move-b}.

@@

TODO: Common scenarios first and then consider more unusual scenarios
that cause cycles or vertices with multiple parents

@@

To start, Alice and Bob both have the graph in \autoref{fig:Move-b}, which
represents the code~\texttt{\hole{} + (\hole{} + y)}.

Alice adds~\vDifferentAlice{} as the left child of \vSimpleTimes{}
and Bob changes \vConstructionY{} to \vDifferentBob{}.
Before transmitting their edits to each other,
Alice and Bob have the graphs in \autoref{fig:Editing Different Parts of the Code-a}
and \autoref{fig:Editing Different Parts of the Code-b}, respectively.
(Note that thee transition edge from \autoref{fig:Move-b} to
\autoref{fig:Editing Different Parts of the Code-b} represents
multiple edges (i.e., deleting~\eConstructionY{} and adding~\eDifferentBob{} along with its child~\vDifferentBob{}).
We thus mark it with a star.)\todo{star size}\todo{"sync"}

@@

Note that aside from transmitting edge deletions and additions,
merging requires no coordination between Alice and Bob.
For example, suppose Alice and Bob did not transmit their edits to each other
but to a third user, Chris, that has the original graph in \autoref{fig:Move-b}.
Chris can apply the edits from Alice and Bob in any order and will always get
the graph in \autoref{fig:Editing Different Parts of the Code-c}.

Stated more formally, our edit model obeys the commutativity shown in \autoref{fig:Commutativity}.
For any sequences of edits from Alice and Bob,
applying them in either order produces the same results.

@@

After Alice and Bob share the edits they made in the previous subsection,
Alice changes~\vDifferentAlice{} to~\vNestedAlice{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-a}.
On the other hand, Bob moves~\vSimpleTimes{} and its children from the left child of~\vWrapPlus{} to its right child
by deleting~\eMoveTimes{} and adding~\eNestedBob{}, which results in the graph in \autoref{fig:Editing Nested Parts of the Code-b}.
Since edits are based on edge identities,
both Alice's and Bob's edits can be applied to the other's graph,
which results in both Alice and Bob having the graph in \autoref{fig:Editing Nested Parts of the Code-c}.

A consequence of this design is that we allow users to modify
parts of the code that have been deleted.  For example,
Alice might receive Bob's deletion of~\eMoveTimes{} before making her
changes and before receiving Bob's addition of~\eNestedBob{}.
Her edits would then temporarily be within the code deleted by Bob.
TODO: \todo{TODO}She can restore that code using a restoration action add a fresh edge to it.

@@

Next we address what happens whn edits conflict.
For example, Alice adds~\vMultiChildAlice{} as the right child of~\vWrapPlus{}
while Bob adds~\vMultiChildBob{} also as the right child of~\vWrapPlus{}.
These result in Alice and Bob having the graphs in \autoref{fig:Multi-child-a} and \autoref{fig:Multi-child-b}, respectively.
These graphs have two different edges for the~\texttt{R} position of~\vWrapPlus{}.
\autoref{fig:Multi-child-a} has~\eMultiChildAlice{}, which points to~\vMultiChildAlice{},
and \autoref{fig:Multi-child-b} has~\eMultiChildBob{}, which points to~\vMultiChildBob{}.
We call this a \emph{multi-child conflict}.

We merge these edits by including \emph{both} edges as edges going out of the \texttt{R} position
of \vWrapPlus{}.
This results in \autoref{fig:Multi-child-c}.
The conflict between Alice's and Bob's changes is
represented by the fact that there are multiple edges going out of one index.
In the user interface, we flag this conflict with the notation~\texttt{\{ x | y \}}
as shown in \autoref{fig:Multi-child-d}.

This conflict must eventually be resolved by the user by deleting
one of the edges and editing the other to be the merged value.
For example, one could delete \eMultiChildBob{} and then change~\vMultiChildAlice{}
to be~\texttt{x * y}.
(Note that users can continue to edit the code while in this conflicted state.
It just must be resolved before the the code is run.\footnote{It
may be possible to develop evaluation models that allow these sorts of conflicts,
but that is beyond the scope of this paper.})

TODO: need example of fully-automatic resolution and semi-automatic resolution.
Note that as a convenience to the user, certain simple
conflicts might be automatically resolved,
but we consider this a higher-level, user-interface consideration.

TODO: Multiple adds cause conflicts (Not usually presented in the UI)

@@

Our tree-based model is similar in that it presents alternate versions
of the code, and it is up to the user to replace those alternates with the desired merged result.
Our model is different in that merge conflicts in the line-based model
prevent some semantic editor services from running
that could continue to run in the tree-based model because
code inside and outside the conflict is still valid code.
It is just the point where they meet the poses a potential problem for editor services.
Another place our model is different
is that changes are explicit rather than being inferred by a \texttt{diff} algorithm.
Finally, in our model, since all changes are in terms of
the structure of the code rather than lines,
all changes respect that structure.
This is unlike the line-based model where there is no guarantee
that a merge conflict follows the grammatical structure of the code.

@@

In addition to conflicts due to multiple edges coming \emph{out of} the same child position of a particular vertex,
conflicts can manifest as multiple edges going \emph{into} a particular vertex.
We call these \emph{multi-parent conflicts}.

An example of this is shown in \autoref{fig:Multi-parent}.
\autoref{fig:Multi-parent-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{(w + \hole{}) + \hole{}}.
Alice moves~\vNestedAlice{} to the~\texttt{R} position of~\vSimpleTimes{},
while Bob moves it to the~\texttt{R} position of~\vWrapPlus{}.
This results in the graphs in \autoref{fig:Multi-parent-b} and \autoref{fig:Multi-parent-c},
which represent respectively the expression~\texttt{(\hole{} + w) + \hole{}} for Alice
and the expression~\texttt{(\hole{} + \hole{}) + w} for Bob.
In both cases, these edits are achieved with one edge delete and one edge add.
Alice deletes~\eNestedAlice{} and adds~\eMultiParentAlice{}, while
Bob deletes~\eNestedAlice{} and adds~\eMultiParentBob{}.
As explained in \autoref{sec:Formalism}, deletes are idempotent so Alice and Bob both deleting~\eNestedAlice{} is not a problem.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob) point to the same vertex.
This generates \autoref{fig:Multi-parent-d}, in which~\vNestedAlice{} has two parents pointing to it.
We call this situation a multi-parent conflict.

@@

TODO: paragraph about why single-user cycles cannot happen.
While single-user edits can be checked to see
whether they create a cycle and thus the system
can immediately warn the user,
multi-user edits may not manifest cycles until after
the edits from multiple users are merged.

For example, consider the situation in \autoref{fig:Multi-User Cycles}.
\autoref{fig:Multi-User Cycles-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{((\hole{} + \hole{}) + (\hole{} * \hole{})) + \hole}.
Alice moves~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath.
This results in \autoref{fig:Multi-User Cycles-b},
which represents the expression~\texttt{(\hole{} + \hole{}) + ((\hole{} * \hole{}) + \hole{})}.
On the other hand, Bob does the same
but puts~\vMultiCycleTimes{} and under~\vMultiCyclePlus{}.
This results in \autoref{fig:Multi-User Cycles-c},
which represents the expression~\texttt{(\hole{} + \hole{}) + ((\hole{} + \hole{}) * \hole{})}.
